Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-03-12 paper/quarto_ecml/paper.tex

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
% \usepackage{amsthm}
% \theoremstyle{plain}
% \newtheorem{corollary}{Research Question}[section]
% \theoremstyle{plain}
% \newtheorem{proposition}{Proposition}[section]
% \theoremstyle{definition}
% \newtheorem{example}{Example}[section]
% \theoremstyle{definition}
% \newtheorem{definition}{Definition}[section]
% \theoremstyle{remark}
% \AtBeginDocument{\renewcommand*{\proofname}{Proof}}
% \newtheorem*{remark}{Remark}
% \newtheorem*{solution}{Solution}
% \newtheorem{refremark}{Remark}[section]
% \newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
  pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
  pdfkeywords={Counterfactual Explanations, Explainable
AI, Representation Learning},
  colorlinks=true,
  linkcolor={black},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
We propose a novel model objective that leverages counterfactual
explanations to increase the explanatory capacity of trained models.
Counterfactual explanations have emerged as a popular tool to explain
predictions made by opaque machine learning models: they inform how
factual inputs would need to change in order for a model to produce some
desired output. Much existing research has focused on generating
counterfactuals that are not only valid but also deemed acceptable with
respect to the stakeholder requirements and plausible with respect to
the underlying data. Recent work has shown that under this premise, the
task of learning plausible explanations---ones that respect the
conditional distribution of the target class---is effectively reassigned
from the model itself to the post-hoc counterfactual explainer. Building
on that work, we propose a novel objective function that employs
counterfactuals (ad-hoc) during the training phase to minimize the
divergence between the learned representations and the explanations.
Through extensive experiments, we demonstrate that our proposed method
facilitates training models that deliver inherently plausible
explanations while maintaining high predictive performance.

\keywords{Counterfactual Explanations \and Explainable
AI \and Representation Learning}

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by advances in \textbf{representation learning}: instead of
relying on features and rules that are carefully hand-crafted by humans,
modern machine learning (ML) models are tasked with learning these
representations from scratch, guided by narrow objectives such as
predictive accuracy \cite{goodfellow2016deep}. Modern advances in
computing have made it possible to provide such models with ever greater
degrees of freedom to achieve that task, which has often led them to
outperform traditionally more parsimonious models. Unfortunately, in
doing so they also learn increasingly complex and highly sensitive
representations that we can no longer easily interpret.

This trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
revolution, \cite{szegedy2013intriguing} showed that artificial neural
networks (ANN) are sensitive to adversarial examples: counterfactuals of
model inputs that yield vastly different model predictions despite being
``imperceptible'' in that they are semantically indifferent from their
factual counterparts. Despite partially effective mitigation strategies
such as \textbf{adversarial training} \cite{goodfellow2014explaining},
truly robust deep learning (DL) remains unattainable even for models
that are considered shallow by today's standards
\cite{kolter2023keynote}.

Part of the problem is that high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives \cite{wilson2020case}\footnote{For clarity: we follow
  standard ML convention in using ``degrees of freedom'' to refer to the
  number of parameters estimated from data.}. Based purely on predictive
performance, these solutions may seem to provide compelling explanations
for the data, when in fact they are based on purely associative,
semantically meaningless patterns. This poses two related challenges:
firstly, it makes these models inherently opaque, since humans cannot
simply interpret what type of explanation the complex learned
representations correspond to; secondly, even if we could resolve the
first challenge, it is not obvious how to mitigate models from learning
representations that correspond to meaningless and implausible
explanations.

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI) which aims to develop tools to derive
explanations from complex model representations. This can mitigate a
scenario in which we deploy opaque models and blindly rely on their
predictions. On countless occasions, this scenario has already occurred
in practice and caused real harm to people who were affected adversely
and often unfairly by automated decision-making systems (ADMS) involving
opaque models \cite{oneil2016weapons}. Effective XAI tools can aide us
in monitoring models and providing recourse to individuals to turn
adverse outcomes (e.g.~``loan application rejected'') into positive ones
(``application accepted''). \cite{wachter2017counterfactual} propose
\textbf{counterfactual explanations} as an effective approach to achieve
this: they explain how factual inputs need to change in order for some
fitted model to produce some desired output, typically involving minimal
perturbations.

To our surprise, the second challenge has not yet attracted any
consolidated research effort. Specifically, there has been no concerted
effort towards improving model \textbf{explainability}, which we define
here as the degree to which learned representations correspond to
explanations that are interpretable and deemed \textbf{plausible} by
humans (see Definition~\ref{def-explainability}). Instead, the choice
has typically been to improve the capacity of XAI tools to identify the
subset explanations that are both plausible and valid for any given
model, independent of whether the learned representations are also
compatible with implausible explanations \cite{altmeyer2024faithful}.
Fortunately, recent findings indicate that explainability can arise as
byproduct of regularization techniques aimed at other objectives such as
robustness, generalization and generative capacity
\cite{schut2021generating}.

Building on these findings, we introduce \textbf{counterfactual
training}: a novel regularization technique geared explicitly towards
aligning model representations with plausible explanations. Our
contributions are as follows:

@@

To the best of our knowledge, our proposed framework for counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed plausible with respect to
the underlying data and stakeholder requirements. To make this more
concrete, we follow \cite{augustin2020adversarial} in tieing the concept
of explainability to the quality of counterfactual explanations that we
can generate for a given model. The authors show that counterfactual
explanations---understood here as minimal input perturbations that yield
some desired model prediction---are generally more meaningful if the
underlying model is more robust to adversarial examples. We can make
intuitive sense of this finding when looking at adversarial training
(AT) through the lens of representation learning with high degrees of
freedom: by inducing models to ``unlearn'' representations that are
susceptible to worst-case counterfactuals (i.e.~adversarial examples),
AT effectively removes some implausible explanations from the solution
space.

@@

This interpretation of the link between explainability through
counterfactuals on one side, and robustness to adversarial examples on
the other, is backed by empirical evidence.
\cite{sauer2021counterfactual} demonstrate that using counterfactual
images during classifier training improves model robustness. Similarly,
\cite{abbasnejad2020counterfactual} argue that counterfactuals represent
potentially useful training data in machine learning, especially in
supervised settings where inputs may be reasonably mapped to multiple
outputs. They, too, demonstrate the augmenting the training data of
image classifiers can improve generalization. \cite{teney2020learning}
propose an approach using counterfactuals in training that does not rely
on data augmentation: they argue that counterfactual pairs typically
already exist in training datasets. Specifically, their approach relies
on, firstly, identifying similar input samples with different
annotations and, secondly, ensuring that the gradient of the classifier
aligns with the vector between pairs of counterfactual inputs using the
cosine distance as a loss function.

@@

There have also been several attempts at formalizing the relationship
between counterfactual explanations (CE) and adversarial examples (AE).
Pointing to clear similarities in how CE and AE are generated,
\cite{freiesleben2022intriguing} makes the case for jointly studying the
opaqueness and robustness problem in representation learning. Formally,
AE can be seen as the subset of CE, for which misclassification is
achieved \cite{freiesleben2022intriguing}. Similarly,
\cite{pawelczyk2022exploring} show that CE and AE are equivalent under
certain conditions and derive theoretical upper bounds on the distances
between them.

Two recent works are closely related to ours in that they use
counterfactuals during training with the explicit goal of affecting
certain properties of post-hoc counterfactual explanations. Firstly,
\cite{ross2021learning} propose a way to train models that are
guaranteed to provide recourse for individuals to move from an adverse
outcome to some positive target class with high probability. The
approach proposed by \cite{ross2021learning} builds on adversarial
training, where in this context susceptibility to targeted adversarial
examples for the positive class is explicitly induced. The proposed
method allows for imposing a set of actionability constraints ex-ante:
for example, users can specify that certain features (e.g.~\emph{age},
\emph{gender}, \ldots) are immutable. Secondly, \cite{guo2023counternet}
are the first to propose an end-to-end training pipeline that includes
counterfactual explanations as part of the training procedure. In
particular, they propose a specific network architecture that includes a
predictor and CE generator network, where the parameters of the CE
generator network are learnable. Counterfactuals are generated during
each training iteration and fed back to the predictor network. In
contrast to \cite{guo2023counternet}, we impose no restrictions on the
neural network architecture at all.

@@

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit objective of aligning representations with plausible
explanations that comply with user requirements. In the context of CE,
plausibility has broadly been defined as the degree to which
counterfactuals comply with the underlying data generating process
\cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
Plausibility is a necessary but insufficient condition for using CE to
provide algorithmic recourse (AR) to individuals affected by opaque
models in practice. This is because for recourse recommendations to be
\textbf{actionable}, they need to not only result in plausible
counterfactuals but also be attainable. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they were 20 years older.
Ignoring all other features, this complies with the definition of
plausibility if 40-year-old individuals are in fact more credit-worthy
on average than young adults. But of course this CE does not qualify for
providing actionable recourse to the applicant since \emph{age} is not a
mutable feature. For our intents and purposes, counterfactual training
aims at improving model explainability by aligning models with
counterfactuals that meet both desiderata, plausibility and
actionability. Formally, we define explainability as follows:

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by \cite{wachter2017counterfactual}. For our
purposes, we let \(T\) vary and consider the counterfactual search as
converged as soon as the predicted probability for the target class has
reached a pre-determined threshold, \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\)\footnote{For
  detailed background information on gradient-based counterfactual
  search and convergence see Section~\ref{sec-app-ce}.}.

To train models with high explainability as defined in
Definition~\ref{def-explainability}, we propose to leverage
counterfactuals in the following objective,

\begin{equation}\phantomsection\label{eq-obj}{
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y})
}\end{equation}

where \(\text{yloss}(\cdot)\) is any conventional classification loss
that induces discriminative performance (e.g.~cross-entropy). The two
additional components in Equation~\ref{eq-obj} are explained in more
detail below. For now, they can be sufficiently described as inducing
explainability directly and indirectly by penalizing: 1) the contrastive
divergence, \(\text{div}(\cdot)\), between counterfactuals
\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, 2) the
adversarial loss, \(\text{advloss}(.)\), with respect to nascent
counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). The tradeoff between
the different components can be governed by adjusting the strengths of
the penalties \(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).

@@

\cite{grathwohl2020your} observe that any classifier can be
re-interpreted as a joint energy-based model (JEM) that learns to
discriminate output classes conditional on inputs and generate inputs.
They show that JEMs can be trained to perform well at both tasks by
directly maximizing the joint log-likelihood factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first factor can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
\cite{grathwohl2020your} minimize the contrastive divergence between
samples drawn from \(p_\theta(\mathbf{x})\) and training observations,
i.e.~samples from \(p(\mathbf{x})\).

A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
tend to do well with respect to the plausibility objective in
Definition~\ref{def-explainability}. If we consider samples drawn from
\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
finding, because the JEM objective effectively minimizes the divergence
between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
To generate samples, \cite{grathwohl2020your} rely on Stochastic
Gradient Langevin Dynamics (SGLD) using an uninformative prior for
initialization. This is where we depart from their methodology: instead
of generating samples through SGLD, we propose using counterfactual
explainers to generate counterfactuals for observed training samples.
Specifically, we have

@@

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. In
particular, we set
\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
where \(y^+\) denotes the index of the target class. We generate samples
\(\mathbf{x}_T^\prime\) by first randomly sampling the target class
\(y^+ \sim p(y)\) and then generating a counterfactual explanation for
that target over \(T\) iterations using a gradient-based counterfactual
generator. This is similar to how conditional sampling is used to draw
from \(p_\theta(\mathbf{x})\) in \cite{grathwohl2020your}.

Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
of observed training samples (positive samples) while at same time
increasing the energy of counterfactuals (negative samples)
\cite{du2019implicit}. As the generated counterfactuals get more
plausible (Definition~\ref{def-explainability}) over the cause of
training, these two opposing effects gradually balance each out
\cite{lippe2024uvadlc}.

@@

Since actionability constraints are binding at test time, we should also
impose them when generating \(\mathbf{x}^\prime\) during each training
iteration to align model representations with user requirements. Through
their effect on \(\mathbf{x}^\prime\), both types of constraints
influence model outcomes through Equation~\ref{eq-div}. Here it is
crucial that we avoid penalizing implausibility that arises due to
mutability constraints. For any mutability-constrained feature \(d\)
this can be achieved by enforcing
\(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever perturbing
\(\mathbf{x}^\prime[d]\) in the direction of \(\mathbf{x}[d]\) would
violate mutability constraints. Specifically, we set
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if

>>>>

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
% \usepackage{amsthm}
% \theoremstyle{plain}
% \newtheorem{proposition}{Proposition}[section]
% \theoremstyle{plain}
% \newtheorem{corollary}{Research Question}[section]
% \theoremstyle{definition}
% \newtheorem{definition}{Definition}[section]
% \theoremstyle{definition}
% \newtheorem{example}{Example}[section]
% \theoremstyle{remark}
% \AtBeginDocument{\renewcommand*{\proofname}{Proof}}
% \newtheorem*{remark}{Remark}
% \newtheorem*{solution}{Solution}
% \newtheorem{refremark}{Remark}[section]
% \newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
  pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
  pdfkeywords={Counterfactual Training, Counterfactual
Explanations, Algorithmic Recourse, Explainable AI, Representation
Learning},
  colorlinks=true,
  linkcolor={black},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
We propose a novel training regime called Counterfactual Training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to
stakeholder requirements. Much existing research has therefore focused
on developing post-hoc methods to generate counterfactuals that meet
these desiderata. In this work, we instead hold models directly
accountable for this desired end goal: Counterfactual Training employs
counterfactuals ad-hoc during the training phase to minimize the
divergence between learned representations and plausible, actionable
explanations. We demonstrate empirically and theoretically that our
proposed method facilitates training models that deliver inherently
desirable explanations while maintaining high predictive performance.

\keywords{Counterfactual Training \and Counterfactual
Explanations \and Algorithmic Recourse \and Explainable
AI \and Representation Learning}

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by advances in \textbf{representation learning}: instead of
relying on features and rules that are carefully hand-crafted by humans,
modern machine learning (ML) models are tasked with learning
representations directly from data, guided by narrow objectives such as
predictive accuracy \cite{goodfellow2016deep}. Modern advances in
computing have made it possible to provide such models with ever-growing
degrees of freedom to achieve that task, which has often led them to
outperform traditionally more parsimonious models. Unfortunately, in
doing so, models learn increasingly complex and highly sensitive
representations that we can no longer easily interpret.

The trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
revolution, \cite{szegedy2013intriguing} showed that artificial neural
networks (ANN) are sensitive to adversarial examples: counterfactuals of
model inputs that yield vastly different model predictions despite being
``imperceptible'' in that they are semantically indifferent from their
factual counterparts. Although some partially effective mitigation
strategies have been proposed, for example \textbf{adversarial training}
\cite{goodfellow2014explaining}, truly robust deep learning (DL) remains
unattainable even for models that are considered shallow by today's
standards \cite{kolter2023keynote}.

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives \cite{wilson2020case}.\footnote{We follow the standard ML
  convention, where ``degrees of freedom'' refer to the number of
  parameters estimated from data.} Indeed, recent work on the so called
``lottery ticket hypothesis'' suggests that modern neural networks can
be pruned by up to 90\% while preserving their predictive performance
\cite{frankle2018lottery} and generalizability \cite{morcos2019success}.
Similarly, \cite{zhang2021understanding} showed that state-of-the-art
neural networks are so expressive that they can fit randomly labeled
data. Thus, looking at the predictive performance, the solutions may
seem to provide compelling explanations for the data, when in fact they
are based on purely associative, semantically meaningless patterns. This
poses two related challenges. Firstly, there is no dependable way to
verify if such complex representations correspond to meaningful and
plausible explanations. Secondly, even if we could resolve the first
challenge, it remains undecided how to ensure that models can
\emph{only} learn valuable explanations.

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI) which aims to develop tools to derive
explanations from complex model representations. This can mitigate a
scenario in which we deploy opaque models and blindly rely on their
predictions. On countless occasions, this scenario has occurred in
practice and caused real harm to people who were affected adversely and
often unfairly by automated decision-making (ADM) systems involving
opaque models \cite{oneil2016weapons,mcgregor2021preventing}. Effective
XAI tools can aid us in monitoring models and providing recourse to
individuals to turn adverse outcomes (e.g., ``loan application
rejected'') into positive ones (e.g., ``application accepted'').
\cite{wachter2017counterfactual} propose \textbf{counterfactual
explanations} (CE) as an effective approach to achieve this goal: CEs
explain how factual inputs need to change in order for some fitted model
to produce some desired output, typically involving minimal
perturbations.

To our surprise, the second challenge has not yet attracted any major
consolidated research effort. Specifically, there has been no concerted
effort towards improving improving models' explanatory capacity, which
we will henceforth simply call ``explainability'', defined as the degree
to which learned representations correspond to explanations that are
interpretable and deemed \textbf{plausible} by humans (see
Definition~\ref{def-explainability}). Instead, the choice has typically
been to improve the ability of XAI tools to identify the subset
explanations that are both plausible and valid for any given model,
independent of whether the learned representations are also compatible
with implausible explanations \cite{altmeyer2024faithful}. Fortunately,
recent findings indicate that explainability can arise as byproduct of
regularization techniques aimed at other objectives such as robustness,
generalization, and generative capacity
\cite{schut2021generating,augustin2020adversarial,altmeyer2024faithful}.

Building on these findings, we introduce \textbf{counterfactual
training}: a novel training regime explicitly geared towards aligning
model representations with plausible explanations. Our contributions are
as follows:

@@

To the best of our knowledge, our proposed framework of counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed plausible with respect to
the underlying data and stakeholder requirements. To make this more
concrete, we follow \cite{augustin2020adversarial} in tying the concept
of explainability to the quality of counterfactual explanations that we
can generate for a given model. The authors show that counterfactual
explanations---understood here as minimal input perturbations that yield
some desired model prediction---are generally more meaningful if the
underlying model is more robust to adversarial examples. We can make
intuitive sense of this finding when looking at adversarial training
(AT) through the lens of representation learning with high degrees of
freedom: by inducing models to ``unlearn'' representations that are
susceptible to worst-case counterfactuals (i.e., adversarial examples),
AT effectively removes some implausible explanations from the solution
space.

@@

This interpretation of the link between explainability through
counterfactuals on one side, and robustness to adversarial examples on
the other, is backed by empirical evidence.
\cite{sauer2021counterfactual} demonstrate that using counterfactual
images during classifier training improves model robustness. Similarly,
\cite{abbasnejad2020counterfactual} argue that counterfactuals represent
potentially useful training data in machine learning, especially in
supervised settings where inputs may be reasonably mapped to multiple
outputs. They, too, demonstrate the augmenting the training data of
image classifiers can improve generalization. \cite{teney2020learning}
propose an approach using counterfactuals in training that does not rely
on data augmentation: they argue that counterfactual pairs typically
already exist in training datasets. Specifically, their approach relies
on identifying similar input samples with different annotations and
ensuring that the gradient of the classifier aligns with the vector
between such pairs of counterfactual inputs using the cosine distance as
the loss function.

@@

There have also been several attempts at formalizing the relationship
between counterfactual explanations and adversarial examples (AE).
Pointing to clear similarities in how CE and AE are generated,
\cite{freiesleben2022intriguing} makes the case for jointly studying the
opaqueness and robustness problem in representation learning. Formally,
AE can be seen as the subset of CE for which misclassification is
achieved \cite{freiesleben2022intriguing}. Similarly,
\cite{pawelczyk2022exploring} show that CE and AE are equivalent under
certain conditions and derive theoretical upper bounds on the distances
between them.

Two recent works are closely related to ours in that they use
counterfactuals during training with the explicit goal of affecting
certain properties of post-hoc counterfactual explanations. Firstly,
\cite{ross2021learning} propose a way to train models that are
guaranteed to provide recourse for individuals to move from an adverse
outcome to some positive target class with high probability. Their
approach builds on adversarial training, where in this context
susceptibility to targeted adversarial examples for the positive class
is explicitly induced. The proposed method allows for imposing a set of
actionability constraints ex-ante: for example, users can specify that
certain features (e.g., \emph{age}, \emph{gender}, \ldots) are
immutable. Secondly, \cite{guo2023counternet} are the first to propose
an end-to-end training pipeline that includes counterfactual
explanations as part of the training procedure. In particular, they
propose a specific network architecture that includes a predictor and CE
generator network, where the parameters of the CE generator network are
learnable. Counterfactuals are generated during each training iteration
and fed back to the predictor network. In contrast to
\cite{guo2023counternet}, we impose no restrictions on the neural
network architecture at all.

@@

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit objective of aligning representations with plausible
explanations that comply with user requirements. In the context of CEs,
plausibility has broadly been defined as the degree to which
counterfactuals comply with the underlying data generating process
\cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
Plausibility is a necessary but insufficient condition for using CEs to
provide algorithmic recourse (AR) to individuals affected by opaque
models in practice. This is because for recourse recommendations to be
\textbf{actionable}, they need to not only result in plausible
counterfactuals but also be attainable. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they were 20 years older.
Ignoring all other features, this would comply with the definition of
plausibility if 40-year-old individuals were in fact more credit-worthy
on average than young adults. But of course this CE does not qualify for
providing actionable recourse to the applicant since \emph{age} is not a
(directly) mutable feature. For our intents and purposes, counterfactual
training aims to improve model explainability by aligning models with
counterfactuals that meet both desiderata, plausibility and
actionability. Formally, we define explainability as follows:

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by \cite{wachter2017counterfactual}. For our
purposes, we let \(T\) vary and consider the counterfactual search as
converged as soon as the predicted probability for the target class has
reached a pre-determined threshold, \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
  background information on gradient-based counterfactual search and
  convergence see supplementary appendix.}

To train models with high explainability as defined in
Definition~\ref{def-explainability}, we propose to leverage
counterfactuals in the following objective:

\begin{equation}\phantomsection\label{eq-obj}{
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
}\end{equation}

where \(\text{yloss}(\cdot)\) is any conventional classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms in Equation~\ref{eq-obj} are explained in more
detail below. For now, they can be sufficiently described as inducing
explainability directly and indirectly by penalizing: (1) the
contrastive divergence, \(\text{div}(\cdot)\), between mature
counterfactuals \(\mathbf{x}_T^\prime\) and observed samples \(x\) and,
(2) the adversarial loss, \(\text{advloss}(.)\), with respect to nascent
counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
regularises the magnitude of the energy terms involved in
\(\text{div}(\cdot)\) \cite{du2019implicit}. The tradeoff between the
different components can be governed by adjusting the strengths of the
penalties \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\).

@@

\cite{grathwohl2020your} observe that any classifier can be
re-interpreted as a joint energy-based model (JEM) that learns to
discriminate output classes conditional on the observed (training)
samples from \(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). They show that JEMs can be trained to perform
well at both tasks by directly maximizing the joint log-likelihood
factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first factor can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
\cite{grathwohl2020your} minimize the contrastive divergence between
these observed samples from \(p(\mathbf{x})\) and generated samples from
\(p_\theta(\mathbf{x})\).

A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
tend to do well with respect to the plausibility objective in
Definition~\ref{def-explainability}. If we consider samples drawn from
\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
finding, because the JEM objective effectively minimizes the divergence
between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
To generate samples, \cite{grathwohl2020your} rely on Stochastic
Gradient Langevin Dynamics (SGLD) using an uninformative prior for
initialization. This is where we depart from their methodology: instead
of SGLD, we propose to use counterfactual explainers to generate
counterfactuals of observed training samples. Specifically, we have:

@@

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. In
particular, we set
\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]\)
where \(y^+\) denotes the index of the randomly drawn target class,
\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed data point
sampled from target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\).
Conditional on the target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes
a mature counterfactual for a randomly sampled factual from a non-target
class generated through a gradient-based counterfactual generator for at
most \(T\) iterations. We define mature counterfactuals as those that
have either exhausted \(T\) or reached convergence in terms of the
pre-determined decision threshold earlier.

Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
of observed training samples (positive samples) while at same time
increasing the energy of counterfactuals (negative samples)
\cite{du2019implicit}. As the generated counterfactuals get more
plausible (Definition~\ref{def-explainability}) over the course of
training, these two opposing effects gradually balance each out
\cite{lippe2024uvadlc}.

@@

Since actionability constraints are binding at test time, we should also
impose them when generating \(\mathbf{x}^\prime\) during each training
iteration to align model representations with user requirements. Through
their effect on \(\mathbf{x}^\prime\), both types of constraints
influence model outcomes through Equation~\ref{eq-div}. Here it is
crucial that we avoid penalizing implausibility that arises due to
mutability constraints. For any mutability-constrained feature \(d\)
this can be achieved by enforcing
\(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever perturbing
\(\mathbf{x}^\prime[d]\) in the direction of \(\mathbf{x}[d]\) would
violate mutability constraints. Specifically, we set
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if:
